Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2021-12-02 Manuscript/Manuscript.tex
First draft introduction section

We talk of multilevel data when there is some kind of hierarchy or
clustering in a dataset. In the typical case, individuals are nested
within groups, but there are many different types of multilevel data. In
the medical field clustering occurs at e.g., the hospitals/center level
in registry data, or at the study-level in meta-analyses (IPDMA). In the
social sciences and official statistics we can find clustering e.g.~at
the country-level, or as imposed by the sampling design. In this paper,
we will refer to the grouping variable as `cluster', and the grouped
variable as `(sample) unit'. For reasons of brevity, we only discuss
clustering between units, not longitudinal data (within-unit
clustering).

Multilevel data requires special care when doing any sort of analysis.
The cluster to which a unit belongs may influence the unit-level
observations, and since clusters are made up of units, clusters depend
on units as well \citep{hox17}. {[}Explain ICC here? The percentage of
variance attributed to the cluster-level is expressed by the intra-class
coefficient (ICC). The ICC can also be interpreted as the expected
correlation between two randomly sampled units in same cluster.{]} These
relations can and should be taken into account when developing analysis
models for multilevel data. At least, such models include a separate
intercept term for each cluster. But there may also be random predictor
effects and/or random error terms (residual error variances), see e.g.
\citet{hox17} and \citet{jong21}. Heterogeneity refers to variability
within clusters vs.~variability between clusters. There are many names
for models that take clustering into account. Some popular examples are
`multilevel models', `hierarchical models', `mixed effect models' and
`random effect models'.

@@

\begin{itemize}
\item
  Why/where does missingness occur in multilevel data? I.e., not only
  patient-level but also cluster-level.
\item
  How can we categorize this? Systematic vs sporadic missingness, see
  \citet{resc13}. Within systematic we have two flavors: unobserved
  constants (same value per cluster) and non-measured random variables
  (which may differ per unit within clusters). In figure 1, we show a
  dataset with units in the rows and variables in the columns, there are
  5 units nested within 2 clusters, and 3 variables of interest.
  Variable \texttt{X1} is completely observed. Variable \texttt{X2} is
  systematically missing, \texttt{X3} is sporadically missing. The
  unobserved value for units 4 and 5 on variable \texttt{X2} may be the
  same or different, defining which type of systematic missingness is
  happening.
\end{itemize}

@@

\begin{itemize}
\item
  What kinds of missingness are there? ADD: missingness mechanisms here.
  See e.g. \citet{yuce08} and \citet{hox15}.
\item
  Why are standard (ad hoc) missing data methods not well suited?
\item
  What types of multilevel methods are available? General overview of
  approaches, see \citet{audi18} and \citet{grun18}. E.g., imputation of
  study level versus patient-level covariates, and one-stage imputation
  versus two-stage imputation methods.
\item
  Additional difficulty that is addressed in this tutorial: MNAR data.
\end{itemize}

@@

\begin{itemize}
\item
  Provide practical guidelines with code snippets for imputation of
  incomplete multilevel data.
\item
  We focus on the workflow for conditional modeling (not JOMO) in
  \texttt{mice}. Refer to other packages: \texttt{mitml},
  \texttt{miceadds}, \texttt{mdmb}.
\item
  Case study options: \texttt{metamisc::impact} (real IPD on traumatic
  brain injuries, without \texttt{NA}s), \texttt{mice::popularity}
  (simulated data on school kids, with MNAR/MAR mixture). TODO: Check
  example data Gelman.
\item
  Introduce case study and set scope of this tutorial: We're providing
  an overview of implementations. It's up-to the reader to decide which
  strategy suits their data. So we won't go into detail for the
  different methods (and equations). This paper is just a software
  tutorial. We'll keep it practical. -\textgreater{} ADD: some kind of
  help function that suggests a suitable predictor matrix to the user,
  given a certain analysis model.
\end{itemize}

@@

We'll use the IMPACT data (\texttt{metamisc::impact}) and a MAR/MNAR
version of the \texttt{mice::popmis} data (i.e., a variation on the Hox
(2010) popularity data, where the missingness in the variables is either
missing at random (MAR) or missing not at random (MNAR)).
-\textgreater{} ask whether we can use the Heckman repo data or simulate
data ourselves

Heckman options:

\begin{itemize}
\item
  leiden85
\item
  GJRM::hiv (\url{https://rdrr.io/github/egeminiani/GJRM/man/hiv.html})
\item
  simulating
\item
  IMPACT
\end{itemize}

\hypertarget{case-study-i-impact}{%
\subsection{Case study I: IMPACT}\label{case-study-i-impact}}

\begin{itemize}
\item
  What does the data look like? \texttt{impact} is traumatic brain
  injury data with patients clustered in studies,
  \(n_{\text{participants}} = 11022\) and \(n_{\text{clusters}} = 15\),
  on the following 11 variables:

  \begin{itemize}
  \tightlist
  \item
    \texttt{name} Name of the study,
  \item
    \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
    observational cohort),
  \item
    \texttt{age} Age of the patient,
  \item
    \texttt{motor\_score} Glasgow Coma Scale motor score,
  \item
    \texttt{pupil} Pupillary reactivity,
  \item
    \texttt{ct} Marshall Computerized Tomography classification,
  \item
    \texttt{hypox} Hypoxia (0=no, 1=yes),
  \item
    \texttt{hypots} Hypotension (0=no, 1=yes),
  \item
    \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
  \item
    \texttt{edh} Epidural hematoma (0=no, 1=yes),
  \item
    \texttt{mort} 6-month mortality (0=alive, 1=dead).
  \end{itemize}
\end{itemize}

-\textgreater{} Why are there no missings? According to the
\href{https://cran.r-project.org/web/packages/metamisc/metamisc.pdf}{vignette},
the data is already imputed (Steyerberg et al, 2008).

\begin{itemize}
\item
  MAR miss varying by cluster. Obs data patt differ per cluster. E.g.,
  in cluster 1 miss depends on age but not in cluster two. Split the
  dataframe and run \texttt{ampute()} on each cluster. -\textgreater{}
  TODO: also make MNAR missingness to heckman model. Maybe based on
  \texttt{ct} variable? Inclusion-selection variable. -\textgreater{}
  otherwise: use \texttt{leiden85} data on blood pressure with MNAR.
  Then run cox regression like the boshuizen article but with living
  situation as clusters. -\textgreater{} TODO: get analyses from
  \url{https://www.gerkovink.com/mimp/Contents/Exercises/Day\%203\%20-\%20Wednesday/Sensitivity_analysis/Sensitivity_analysis.html}.
\item
  ADD: \texttt{multilevel\_ampute()} wrapper function in \texttt{mice}.
\end{itemize}

\hypertarget{case-study-ii-popularity}{%
\subsection{Case Study II: Popularity}\label{case-study-ii-popularity}}

\begin{itemize}
\item
  What does the data look like? \texttt{popNCR} is a simulated dataset
  with pupils clustered in classes, \(n_{\text{participants}} = 2000\),
  \(n_{\text{clusters}} = 100\), on 7 variables:

  \begin{itemize}
  \tightlist
  \item
    \texttt{pupil} Pupil number within class,
  \item
    \texttt{class} Class number,
  \item
    \texttt{extrav} Pupil extraversion,
  \item
    \texttt{sex} Pupil gender,
  \item
    \texttt{texp} Teacher experience (years),
  \item
    \texttt{popular} Pupil popularity,
  \item
    \texttt{popteach} Teacher popularity.
  \end{itemize}
\item
  What are the ICCs? For \texttt{popular} the ICC is 0.33. For
  \texttt{popteach} it is 0.31. It would be wise to use multilevel
  modeling.
\item
  What does the missingness look like? Induced MAR/MNAR missingness.
  Missing data pattern:
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Does the missing data of \texttt{popular} depend on \texttt{popteach}?
  Does the missingness in teacher popularity depend on pupil popularity?
  -\textgreater{} Check this by making a histogram of \texttt{popteach}
  separately for the pupils with known popularity and missing
  popularity, and the other way around.
\end{itemize}

@@

\begin{itemize}
\item
  We do see that the distribution for the missing \texttt{popular} is
  further to the right than the distribution for observed
  \texttt{popular}. This would indicate a right-tailed MAR missingness.
  In fact this is exactly what happens, because we created the
  missingness in these data ourselves. But we made it observable by
  examining the relations between the missingness in popular and the
  observed data in \texttt{popteach}. There is also a dependency between
  the missingness in teacher popularity and pupil popularity. The
  relation seems to be right-tailed as well.
\item
  We can impute the missingness the `standard' way, ignoring the
  multilevel structure of the data. This is surely invalid, given the
  high ICCs, but we'll do it anyways.
\item
  We'll use predictive mean matching to impute the continuous variables
  (some appear to be somewhat ordinal), and logistic regression to
  impute the binary variable \texttt{sex}. We do not use the observation
  identifier \texttt{pupil} or cluster identifier \texttt{class} as
  predictors to impute other variables.
\end{itemize}

@@

\begin{CodeInput}
R> # compare descriptives before and after imputation
R> psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65   11.0   1  26.0  5.97
class*   2000 50.37   51.0   1 100.0 29.08
extrav   1484  5.31    5.0   1  10.0  1.29
sex*     1504  1.56    2.0   1   2.0  0.50
texp     1024 11.80   12.0   2  25.0  6.26
popular  1490  4.83    4.8   0   9.1  1.34
popteach 1472  4.83    5.0   1  10.0  1.36
school   2000  5.54    6.0   1  10.0  2.89
\end{CodeOutput}
\begin{CodeInput}
R> psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65     11   1  26.0  5.97
class*   2000 50.37     51   1 100.0 29.08
extrav   2000  5.25      5   1  10.0  1.27
sex*     2000  1.53      2   1   2.0  0.50
texp     2000 12.48     12   2  25.0  6.29
popular  2000  4.99      5   0   9.1  1.35
popteach 2000  5.02      5   1  10.0  1.39
school   2000  5.54      6   1  10.0  2.89
\end{CodeOutput}
\begin{CodeInput}
R> # TODO: add stripplot with boxplot overlay instead of the tables (make pooled one thick on top)
R> # TODO: pool mean median and sd
R> # feedback Stef: numbers, continuous statistics such as means, and uncertainty estimates. So we can pool the sd's. And leave out the min and max, because those are not normally distr.
R> # TODO: add FMI for each of the estimates? at least for the mean
R> 
R> # further inspection of the imputations
R> densityplot(imp_ignored)
\end{CodeInput}

@@

\begin{CodeInput}
R> # compare ICCs before and after imputation
R> ICCs <- data.frame(
+   vars = c("popular", "popteach", "texp"), 
+   incomplete = c(multilevel::ICC1(aov(popular ~ class, pop)), 
+                multilevel::ICC1(aov(popteach ~ class, pop)),
+                multilevel::ICC1(aov(texp ~ class, pop))), 
+   ignored = c(multilevel::ICC1(aov(popular ~ class, complete(imp_ignored))), 
+               multilevel::ICC1(aov(popteach ~ class, complete(imp_ignored))), 
+               multilevel::ICC1(aov(texp ~ class, complete(imp_ignored))))
+   )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored
1  popular  0.3280070 0.2716802
2 popteach  0.3138658 0.2528468
3     texp  1.0000000 0.4395296
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # adjust the predictor matrix
R> pred <- ini$pred 
R> pred[, "pupil"] <- 0
R> pred
\end{CodeInput}
\begin{CodeOutput}
         pupil class extrav sex texp popular popteach school
pupil        0     1      1   1    1       1        1      1
class        0     0      1   1    1       1        1      1
extrav       0     1      0   1    1       1        1      1
sex          0     1      1   0    1       1        1      1
texp         0     1      1   1    0       1        1      1
popular      0     1      1   1    1       0        1      1
popteach     0     1      1   1    1       1        0      1
school       0     1      1   1    1       1        1      0
\end{CodeOutput}
\begin{CodeInput}
R> # impute the data, cluster as predictor
R> imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)
\end{CodeInput}
\begin{CodeOutput}
Warning: Number of logged events: 325
\end{CodeOutput}
\begin{CodeInput}
R> # check logged events
R> head(imp_predictor$loggedEvents)
\end{CodeInput}
\begin{CodeOutput}
  it im      dep   meth
1  1  1   extrav    pmm
2  1  1      sex logreg
3  1  1     texp    pmm
4  1  1  popular    pmm
5  1  1 popteach    pmm
6  1  1 popteach    pmm
                                                                                                                                                                                                                                                       out
1                                                                                                                                                                                                                                                   school
2                                                                                                                                                                                                                                                   school
3                                                                                                                                                                                                                                                   school
4                                                                                                                                                                                                                                             texp, school
5                                                                                                                                                                                                                                                     texp
6 mice detected that your data are (nearly) multi-collinear.\nIt applied a ridge penalty to continue calculations, but the results can be unstable.\nDoes your dataset contain duplicates, linear transformation, or factors with unique respondent names?
\end{CodeOutput}
\begin{CodeInput}
R> ## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and popteach.
R> 
R> # check convergence of the imputation model
R> plot(imp_predictor)
\end{CodeInput}

@@

\begin{CodeInput}
R> # compare descriptives before and after imputation
R> psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65   11.0   1  26.0  5.97
class*   2000 50.37   51.0   1 100.0 29.08
extrav   1484  5.31    5.0   1  10.0  1.29
sex*     1504  1.56    2.0   1   2.0  0.50
texp     1024 11.80   12.0   2  25.0  6.26
popular  1490  4.83    4.8   0   9.1  1.34
popteach 1472  4.83    5.0   1  10.0  1.36
school   2000  5.54    6.0   1  10.0  2.89
\end{CodeOutput}
\begin{CodeInput}
R> psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65     11   1  26.0  5.97
class*   2000 50.37     51   1 100.0 29.08
extrav   2000  6.20      6   1  10.0  2.00
sex*     2000  1.52      2   1   2.0  0.50
texp     2000 14.26     15   2  25.0  6.55
popular  2000  5.03      5   0   9.1  1.35
popteach 2000  5.03      5   1  10.0  1.38
school   2000  5.54      6   1  10.0  2.89
\end{CodeOutput}
\begin{CodeInput}
R> # further inspection of the imputations
R> densityplot(imp_predictor)
\end{CodeInput}

@@

\begin{CodeInput}
R> # compare ICCs before and after imputation
R> ICCs <- ICCs %>% cbind(
+            predictor = c(multilevel::ICC1(aov(popular ~ class, complete(imp_predictor))), 
+                         multilevel::ICC1(aov(popteach ~ class, complete(imp_predictor))), 
+                         multilevel::ICC1(aov(texp ~ class, complete(imp_predictor))))
+            )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored predictor
1  popular  0.3280070 0.2716802 0.3550567
2 popteach  0.3138658 0.2528468 0.3306035
3     texp  1.0000000 0.4395296 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\hypertarget{amputation}{%
\subsection{Amputation}\label{amputation}}

\begin{itemize}
\tightlist
\item
\end{itemize}

\hypertarget{modeling-choices}{%
\subsection{Modeling choices}\label{modeling-choices}}

\begin{itemize}
\item
  Which models will we discuss? We'll build the model to grow in
  complexity. The final model is the most complex but also the most
  versatile.
\item
  Note on model complexity: Typically, we should at least use random
  intercepts, but often random slopes as well. Ideally we impute with
  random everything and heteroscedastic errors: most generic method (no
  worry about congeniality, but don't mention the term) -\textgreater{}
  Refer to other papers for background, we'll focus just on the software
  implementation of the situations mentioned there. Sometimes there's
  little reason to assume some variable is affected by heterogeneity.
  -\textgreater{} Refer to \citet{meng94}, an Audigier paper, and a
  paper by Grund on congeniality and random slopes.
\item
  Step 0: As predictor + CCA to scare off users
\item
  Step 1: Random intercepts
\item
  Step 2: Random slopes
\item
  Step 3: Residuals
\item
  Heckman model for MNAR
\item
  What do the different implementations look like? How to define the
  imputation model(s) in \texttt{mice}?
\end{itemize}

\hypertarget{step-0}{%
\subsection{Step 0}\label{step-0}}

\begin{itemize}
\item
  AKA multilevel imputation for dummies.
\item
  Doesn't work for systematic missingness.
\end{itemize}

\hypertarget{step-1-3-mnar}{%
\subsection{Step 1-3 + MNAR}\label{step-1-3-mnar}}

\begin{itemize}
\tightlist
\item
  TODO: fill in.
\end{itemize}

@@

\begin{itemize}
\item
  Analysis of scientific interest.
\item
  Pooling using \texttt{mitml}.
\item
  Pooling `regular' parameters vs more `exotic' parameters (SE of
  residual errors, or autocorrelation)
\item
  ADD: export \texttt{mids} objects to other packages like \texttt{lme4}
  or \texttt{coxme}?
\end{itemize}

@@

\begin{itemize}
\item
  JOMO in \texttt{mice} --\textgreater{} on the side for now
\item
  Additional levels of clustering
\item
  More complex data types: timeseries and polynomial relationship in the
  clustering.
\end{itemize}

>>>>

We talk of multilevel data when there is some kind of hierarchy or
clustering in a dataset. In the typical case, individuals are nested
within groups, but there are many different types of multilevel data. In
the medical field, clustering occurs at e.g., the hospitals/center level
in registry data, or at the study-level in meta-analyses (IPDMA). In the
social sciences and official statistics we can find clustering e.g.~at
the country-level, or as imposed by the sampling design. In this paper,
we will refer to the grouping variable as `cluster', and the grouped
variable as `(sample) unit'.\footnote{Add that we'll only discuss two
  levels?} For reasons of brevity, we only discuss clustering between
units, not within units (such as in timeseries or longitudinal data).

Analyzing multilevel data requires special care, compared to `regular',
single level data. The cluster to which a unit belongs may influence the
unit-level observations, and since clusters are made up of units,
clusters depend on units as well \citep{hox17}. These relations can and
should be taken into account when developing analysis models for
multilevel data.\footnote{Explain ICC here? The percentage of variance
  attributed to the cluster-level is expressed by the intra-class
  coefficient (ICC). The ICC can also be interpreted as the expected
  correlation between two randomly sampled units in same cluster. So if
  the ICC is high, a lot of variability in a variable is due to the
  clustering, which should be modeled accordingly.} Multilevel models
typically include separate intercepts for each cluster, which relieves
one restriction imposed by single-level models: equal group means across
clusters. Additionally, there may be random predictor effects and/or
random error terms (residual error variances), see e.g. \citet{hox17}
and \citet{jong21}.\footnote{Add that heterogeneity refers to
  variability within clusters.} There are many names for models that
take clustering into account. Some popular examples are `multilevel
models', `hierarchical models', `mixed effect models' and `random effect
models'.

@@

Multilevel data is not spared the ubiquitous problem of missing
information. Just as in single level data, missingness may occur at the
unit level. But with multiple levels of data comes the potential for
missingness at multiple levels. Missingness in multilevel data can be
categorized into two general patterns: systematical missingness and
sporadic missingness, see \citet{resc13}. In figure 1, we show a dataset
with units in the rows and variables in the columns, there are 5 units
nested within 2 clusters, and 3 variables of interest. Variable
\texttt{X1} is completely observed. Variable \texttt{X2} is
systematically missing, \texttt{X3} is sporadically missing.\footnote{Explain
  why.} Systematic missingness can be further subdivided into unobserved
constants (same value per cluster) and non-measured random variables
(which may differ per unit within clusters). In Figure 1, the former
implies that the unobserved values for units 4 and 5 on variable
\texttt{X2} would be equal. With the latter, the values may differ.
Depending on the missing data pattern, there are more or less optimal
way of accommodating the missingness.

Ignoring missing data in research endeavors is almost never a good idea.
Complete case analysis (i.e., excluding all units with one or more
missing entries) can introduce bias in statistical inference and lowers
statistical power. Instead, the missingness should be accommodated
\emph{before} or \emph{within} the analysis of scientific interest.
Especially the former is very generic and popular. Imputing (i.e.,
filling in) the missing values splits the missing data problem from the
scientific problem. The \proglang{R} package \pkg{mice} has become the
de-facto standard for imputation by chained equations, which solves the
missingness one variable at a time, iteratively. \pkg{mice} is known to
yield valid inferences under many different missing data circumstances
\citep{buur18}. In this paper, we'll discuss how to use \pkg{mice} in
the context of multilevel data, under varying missing data
mechanisms.\footnote{Discuss missingness mechanisms before this point,
  add references \citet{yuce08} and \citet{hox15}.}

@@

@@

This papers serves as a tutorial for imputing incomplete multilevel data
with \pkg{mice}. We provide practical guidelines and code snippets for
different missing data situations. For reasons of brevity, we focus on
imputation by chained equations, although JOMO is available in
\pkg{mice} as well. Other useful packages for incomplete multilevel data
include: \pkg{mitml}, \pkg{miceadds}, \pkg{mdmb}.\footnote{Rephrase:
  Some level of knowledge on multilevel models is assumed. We're
  providing an overview of implementations. It's up-to the reader to
  decide which multilevel strategy suits their data. So we won't go into
  detail for the different methods (and equations). Refer to
  \citet{meng94}, an Audigier paper, and a paper by Grund on
  congeniality and random slopes. This paper is just a software
  tutorial. We'll keep it practical.}

We structure this tutorial around three case studies:

\begin{itemize}
\item
  \texttt{mice::popmis} (simulated data on school kids, with MNAR/MAR
  mixture);
\item
  \texttt{metamisc::impact} (real IPD on traumatic brain injuries,
  without \texttt{NA}s);
\item
  \texttt{GJRM::hiv} (simulated patient data on HIV, without
  \texttt{NA}s)
\end{itemize}

For each case study we focus on a different aspect to illustrate how to
impute incomplete multilevel data.

@@

We introduce three case studies to illustrate the workflow. In the
\texttt{mice::popmis} data, we show the advantages of including the
multilevel structure of the data into the imputation model. In the
\texttt{metamisc::impact} data we'll show how to induce missingness and
solve it in real-world data. In the \texttt{GJRM::hiv} we provide novel
methodology\footnote{not really, the methods exist already, but how to
  show that this is something new and exciting?} for imputing MNAR
missingness according to the Heckman model.

For each case study we'll look at least at: 1) the incomplete data; 2)
the imputation model; 3) the imputed data; and 4) how to obtain pooled
estimates for the analysis of scientific interest.

\hypertarget{case-study-i-popularity}{%
\subsection{Case Study I: Popularity}\label{case-study-i-popularity}}

\texttt{popNCR} is a simulated dataset with pupils clustered in classes,
\(n_{\text{participants}} = 2000\), \(n_{\text{clusters}} = 100\), on 7
variables:

\begin{itemize}
\tightlist
\item
  \texttt{pupil} Pupil number within class,
\item
  \texttt{class} Class number,
\item
  \texttt{extrav} Pupil extraversion,
\item
  \texttt{sex} Pupil gender,
\item
  \texttt{texp} Teacher experience (years),
\item
  \texttt{popular} Pupil popularity,
\item
  \texttt{popteach} Teacher popularity.
\end{itemize}

\hypertarget{the-incomplete-data}{%
\subsubsection{The incomplete data}\label{the-incomplete-data}}

\begin{itemize}
\tightlist
\item
  What does the missingness look like? Induced MAR/MNAR missingness.
  Missing data pattern:
\end{itemize}

@@

\begin{itemize}
\item
  What are the ICCs? For \texttt{popular} the ICC is 0.33. For
  \texttt{popteach} it is 0.31. It would be wise to use multilevel
  modeling.
\item
  Does the missing data of \texttt{popular} depend on \texttt{popteach}?
  Does the missingness in teacher popularity depend on pupil popularity?
  -\textgreater{} Check this by making a histogram of \texttt{popteach}
  separately for the pupils with known popularity and missing
  popularity, and the other way around.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  We do see that the distribution for the missing \texttt{popular} is
  further to the right than the distribution for observed
  \texttt{popular}. This would indicate a right-tailed MAR missingness.
  In fact this is exactly what happens, because we created the
  missingness in these data ourselves. But we made it observable by
  examining the relations between the missingness in popular and the
  observed data in \texttt{popteach}. There is also a dependency between
  the missingness in teacher popularity and pupil popularity. The
  relation seems to be right-tailed as well.
\end{itemize}

\hypertarget{the-imputation-model}{%
\subsubsection{The imputation model}\label{the-imputation-model}}

\begin{itemize}
\item
  We could ignore the multilevel structure of the data. This is likely
  invalid, given the high ICCs, but we'll do it anyways.
\item
  We'll use predictive mean matching to impute the continuous variables
  (some appear to be somewhat ordinal), and logistic regression to
  impute the binary variable \texttt{sex}. We do not use the observation
  identifier \texttt{pupil} or cluster identifier \texttt{class} as
  predictors to impute other variables.
\end{itemize}

@@

\begin{CodeInput}
R> # compare descriptives before and after imputation
R> psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65   11.0   1  26.0  5.97
class*   2000 50.37   51.0   1 100.0 29.08
extrav   1484  5.31    5.0   1  10.0  1.29
sex*     1504  1.56    2.0   1   2.0  0.50
texp     1024 11.80   12.0   2  25.0  6.26
popular  1490  4.83    4.8   0   9.1  1.34
popteach 1472  4.83    5.0   1  10.0  1.36
school   2000  5.54    6.0   1  10.0  2.89
\end{CodeOutput}
\begin{CodeInput}
R> psych::describe(mice::complete(imp_ignored))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65     11   1  26.0  5.97
class*   2000 50.37     51   1 100.0 29.08
extrav   2000  5.23      5   1  10.0  1.25
sex*     2000  1.51      2   1   2.0  0.50
texp     2000 12.55     12   2  25.0  6.27
popular  2000  4.99      5   0   9.1  1.32
popteach 2000  5.01      5   1  10.0  1.38
school   2000  5.54      6   1  10.0  2.89
\end{CodeOutput}
\begin{CodeInput}
R> # TODO: add stripplot with boxplot overlay instead of the tables (make pooled one thick on top)
R> # TODO: pool mean median and sd
R> # feedback Stef: numbers, continuous statistics such as means, and uncertainty estimates. So we can pool the sd's. And leave out the min and max, because those are not normally distr.
R> # TODO: add FMI for each of the estimates? at least for the mean
R> 
R> # further inspection of the imputations
R> densityplot(imp_ignored)
\end{CodeInput}

@@

\begin{CodeInput}
R> # compare ICCs before and after imputation
R> ICCs <- data.frame(
+   vars = c("popular", "popteach", "texp"), 
+   incomplete = c(multilevel::ICC1(aov(popular ~ class, pop)), 
+                multilevel::ICC1(aov(popteach ~ class, pop)),
+                multilevel::ICC1(aov(texp ~ class, pop))), 
+   ignored = c(multilevel::ICC1(aov(popular ~ class, complete(imp_ignored))), 
+               multilevel::ICC1(aov(popteach ~ class, complete(imp_ignored))), 
+               multilevel::ICC1(aov(texp ~ class, complete(imp_ignored))))
+   )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored
1  popular  0.3280070 0.2754195
2 popteach  0.3138658 0.2467415
3     texp  1.0000000 0.4443760
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # adjust the predictor matrix
R> pred <- ini$pred 
R> pred[, "pupil"] <- 0
R> pred
\end{CodeInput}
\begin{CodeOutput}
         pupil class extrav sex texp popular popteach school
pupil        0     1      1   1    1       1        1      1
class        0     0      1   1    1       1        1      1
extrav       0     1      0   1    1       1        1      1
sex          0     1      1   0    1       1        1      1
texp         0     1      1   1    0       1        1      1
popular      0     1      1   1    1       0        1      1
popteach     0     1      1   1    1       1        0      1
school       0     1      1   1    1       1        1      0
\end{CodeOutput}
\begin{CodeInput}
R> # impute the data, cluster as predictor
R> imp_predictor <- mice(pop, maxit = 10, pred = pred, print = FALSE)
\end{CodeInput}
\begin{CodeOutput}
Warning: Number of logged events: 312
\end{CodeOutput}
\begin{CodeInput}
R> # check logged events
R> head(imp_predictor$loggedEvents)
\end{CodeInput}
\begin{CodeOutput}
  it im      dep   meth
1  1  1   extrav    pmm
2  1  1      sex logreg
3  1  1     texp    pmm
4  1  1  popular    pmm
5  1  1  popular    pmm
6  1  1 popteach    pmm
                                                                                                                                                                                                                                                       out
1                                                                                                                                                                                                                                                   school
2                                                                                                                                                                                                                                                   school
3                                                                                                                                                                                                                                                   school
4                                                                                                                                                                                                                                                     texp
5 mice detected that your data are (nearly) multi-collinear.\nIt applied a ridge penalty to continue calculations, but the results can be unstable.\nDoes your dataset contain duplicates, linear transformation, or factors with unique respondent names?
6                                                                                                                                                                                                                                             texp, school
\end{CodeOutput}
\begin{CodeInput}
R> ## "The mice() function detects multicollinearity, and solves the problem by removing one or more predictors for the model", in this case texp is removed as predictor of popular and popteach.
R> 
R> # check convergence of the imputation model
R> plot(imp_predictor)
\end{CodeInput}

@@

\begin{CodeInput}
R> # compare descriptives before and after imputation
R> psych::describe(pop)[, c("n", "mean", "median", "min", "max", "sd")]
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65   11.0   1  26.0  5.97
class*   2000 50.37   51.0   1 100.0 29.08
extrav   1484  5.31    5.0   1  10.0  1.29
sex*     1504  1.56    2.0   1   2.0  0.50
texp     1024 11.80   12.0   2  25.0  6.26
popular  1490  4.83    4.8   0   9.1  1.34
popteach 1472  4.83    5.0   1  10.0  1.36
school   2000  5.54    6.0   1  10.0  2.89
\end{CodeOutput}
\begin{CodeInput}
R> psych::describe(mice::complete(imp_predictor))[, c("n", "mean", "median", "min", "max", "sd")] #note that this is just 1 imputation, not the pooled results
\end{CodeInput}
\begin{CodeOutput}
            n  mean median min   max    sd
pupil    2000 10.65   11.0   1  26.0  5.97
class*   2000 50.37   51.0   1 100.0 29.08
extrav   2000  5.22    5.0   1  10.0  1.29
sex*     2000  1.53    2.0   1   2.0  0.50
texp     2000 14.26   15.0   2  25.0  6.55
popular  2000  5.59    5.4   0   9.1  1.81
popteach 2000  5.03    5.0   1  10.0  1.40
school   2000  5.54    6.0   1  10.0  2.89
\end{CodeOutput}
\begin{CodeInput}
R> # further inspection of the imputations
R> densityplot(imp_predictor)
\end{CodeInput}

@@

\begin{CodeInput}
R> # compare ICCs before and after imputation
R> ICCs <- ICCs %>% cbind(
+            predictor = c(multilevel::ICC1(aov(popular ~ class, complete(imp_predictor))), 
+                         multilevel::ICC1(aov(popteach ~ class, complete(imp_predictor))), 
+                         multilevel::ICC1(aov(texp ~ class, complete(imp_predictor))))
+            )
R> ICCs
\end{CodeInput}
\begin{CodeOutput}
      vars incomplete   ignored predictor
1  popular  0.3280070 0.2754195 0.2496278
2 popteach  0.3138658 0.2467415 0.3483987
3     texp  1.0000000 0.4443760 1.0000000
\end{CodeOutput}
\end{CodeChunk}

@@

\hypertarget{case-study-ii-impact}{%
\subsection{Case study II: IMPACT}\label{case-study-ii-impact}}

\texttt{impact} is traumatic brain injury data with patients clustered
in studies, \(n_{\text{participants}} = 11022\) and
\(n_{\text{clusters}} = 15\), on the following 11 variables:

\begin{itemize}
\tightlist
\item
  \texttt{name} Name of the study,
\item
  \texttt{type} Type of study (RCT: randomized controlled trial, OBS:
  observational cohort),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{motor\_score} Glasgow Coma Scale motor score,
\item
  \texttt{pupil} Pupillary reactivity,
\item
  \texttt{ct} Marshall Computerized Tomography classification,
\item
  \texttt{hypox} Hypoxia (0=no, 1=yes),
\item
  \texttt{hypots} Hypotension (0=no, 1=yes),
\item
  \texttt{tsah} Traumatic subarachnoid hemorrhage (0=no, 1=yes),
\item
  \texttt{edh} Epidural hematoma (0=no, 1=yes),
\item
  \texttt{mort} 6-month mortality (0=alive, 1=dead).
\end{itemize}

The data is already imputed (Steyerberg et al, 2008), so we'll induce
missingness ourselves. For example, MAR missingness varying by
cluster.\footnote{Observed data pattern should differ per cluster. So,
  in cluster 1, the missingness would depend on age, but not in cluster
  two. Split the dataframe and run \texttt{ampute()} on each cluster.}

@@

\begin{itemize}
\item
  Analysis of scientific interest.
\item
  Pooling using \texttt{mitml}.
\item
  Pooling `regular' parameters vs more `exotic' parameters (SE of
  residual errors, or autocorrelation)
\end{itemize}

@@

\begin{itemize}
\item
  JOMO in \pkg{mice} -\textgreater{} on the side for now
\item
  Additional levels of clustering
\item
  More complex data types: timeseries and polynomial relationship in the
  clustering.
\end{itemize}

\hypertarget{think-about}{%
\section{Think about}\label{think-about}}

\begin{itemize}
\item
  Adding some kind of help function to mice that suggests a suitable
  predictor matrix to the user, given a certain analysis model.
\item
  Adding a \texttt{multilevel\_ampute()} wrapper function in mice.
\item
  Exporting \texttt{mids} objects to other packages like \texttt{lme4}
  or \texttt{coxme}?
\end{itemize}
